---
title: "R Practical Codes"
author: "Dr. Odoi and Seth Antanah"
date: "`r Sys.Date()`"
output: html_document
---

# Practical Codes for Analysis in R

```{r warning=FALSE}

install.packages("summarytools")
install.packages("tidyverse")
install.packages("ggplot2")
```

```{r}

library(summarytools)
library(tidyverse)
library(corrplot)
library(ggplot2)
```

## Exploratory Statistics

Exploratory Statistics refers to the use of statistical techniques to summarize and describe a dataset in order to uncover patterns, relationships, and initial insights. This can involve calculating summary statistics, creating visualizations, and using basic statistical methods to get a sense of the data's characteristics. Its is an important step in statistical data analysis.

> The examples will make illustrations using both functions native in R and a package when necessary. This is done to show the options you have when performing analysis in R.

### Load Data for Exploration

environmental_data.csv is a mocked up data set hence does not contain real life data.

```{r}

environmental_data <- read.csv('environmental_data.csv')
attach(environmental_data)
head(environmental_data)
```

### **Summary Statistics**

Use the simpler **`summary()`** function or the advance **`descr()`** function from the summarytools package, to produce statistical summaries such as the mean, median, min, max, and quartiles..

```{r}

## Commented codes can be uncommented by removing the # symbols
# summary(environmental_data)
summarytools::descr(environmental_data)
```

### Correlation

Calculates correlation coefficients between numeric variables

```{r}

# Correlation between two variables in the dataset
cor(Water_Usage, Pollution_Level)
```

```{r}

# Correlation between all variables in the dataset
numeric_columns <- sapply(environmental_data, is.numeric)
correlation_matrix <- cor(environmental_data[, numeric_columns])

print(correlation_matrix)
```

### Graphics

#### **Histograms**

**`hist()`**: Creates a histogram to visualize the distribution of numeric variables.

```{r}

# Simple Approch
hist(Energy_Consumed, col = "lightgray", border = NULL,
     main = paste("Histogram of" , "Energy Consumed"),
     xlab = "Energy Consumed", ylab = "Frequency",
     axes = TRUE, plot = TRUE, labels = FALSE)
```

```{r}
## Using ggplot2 
gg <- ggplot(environmental_data, aes(x = Pollution_Level)) +
  geom_histogram(binwidth = 10, fill = "gray", color = "black") +
  labs(title = "Histogram of Pollution Levels", x = "Pollution Level")

gg + theme_minimal()
```

```{r}

# Bar plot of Technology Usage
technology_counts <- environmental_data %>%
  group_by(Technology_Used) %>%
  summarise(count = n())
  
gg <- ggplot(technology_counts, aes(x = reorder(Technology_Used, -count), y = count)) +
  geom_bar(stat = "identity", fill = "orange") +
  labs(title = "Bar Plot of Technology Usage", x = "Technology Used", y = "Count") 

gg + theme_minimal()
```

#### **Scatter Plots**

**`plot()`**: Creates scatter plots to visualize the relationship between two numeric variables.

```{r}
plot(x = Pollution_Level, y = Energy_Consumed)
```

```{r}

gg <- ggplot(environmental_data, aes(x = Pollution_Level, y = Energy_Consumed)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of Pollution Level vs. Water Usage", x = "Pollution                   Level", y = "Water Usage")

gg + theme_minimal()
```

#### **Box Plots**

**`plot()`**: Generates a box plot to visualize the distribution and identify outliers in numeric data.

```{r}
plot(x = as.factor(Location), y = Energy_Consumed)
```

```{r}

# Box plot of Energy Consumed by Location
ggplot(environmental_data, aes(x = Location, y = Energy_Consumed)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  labs(title = "Box Plot of Energy Consumed by Location", x = "Location", y = "Energy Consumed")
```

**Correlation Analysis**:

-   **`cor()`**: Calculates correlation coefficients between numeric variables

```{r}

numeric_columns <- sapply(environmental_data, is.numeric)
correlation_matrix <- cor(environmental_data[, numeric_columns])

# Create the correlation plot
corrplot(correlation_matrix, method = "color", type = "upper", tl.col = "black")
```

**Density Plots**:

**`density()`**: Creates a kernel density plot to visualize the probability density of numeric

```{r}

density_plot <- density(Pollution_Level)
plot(density_plot)
```

```{r}


gg <- ggplot(environmental_data, aes(x = Pollution_Level)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Pollution Levels", x = "Pollution Level", y = "Density")

gg + theme_minimal()
```

**Crosstabulations (Contingency Tables)**:

**`table()`**: Creates contingency tables to examine the relationships between two categorical variables.

```{r}

location <- Location
technology <- Technology_Used

cross_tab <- table(location, technology)
print(cross_tab)
```

```{r}

# Create a contingency table
contingency_table <- table(environmental_data$Location, environmental_data$Technology_Used)

# Convert the table to a data frame
contingency_df <- as.data.frame.matrix(contingency_table)

# Create a grouped bar plot
ggplot(contingency_df, aes(x = rownames(contingency_df))) +
  geom_bar(aes(y = `Bioremediation`, fill = "Bioremediation"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = `Solar Power`, fill = "Solar Power"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = `Waste-to-Energy`, fill = "Waste-to-Energy"), stat = "identity", position = "dodge") +
  labs(title = "Technology Usage by Location", x = "Location", y = "Count", fill = "Technology Used") +
  scale_fill_manual(values = c("Bioremediation" = "blue", "Solar Power" = "green", "Waste-to-Energy" = "orange")) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Pairwise Plots**:

**`pairs()`**: Generates a matrix of scatter plots for multiple numeric variables to explore their relationships.

```{r}

data <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c(3, 6, 4, 8, 7),
  z = c(5, 2, 6, 3, 1)
)
pairs(data)
```
